---
title: "AcqVA Aurora Guide to Processing Eye-Tracking Data in R"
author: "Martin Schweinberger"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
bibliography: bibliography.bib
link-citations: yes
---

# Introduction{-}

This document contains a shortened version of the processing chain that was shown in the workshop *Processing and Analyzing Eye-Tracking Data in R* in the AcqVA Aurora Center at the Arctic University of Norway, Tromsø. The R Markdown document for the tutorial can be downloaded [here](https://slcladal.github.io/etpro.Rmd) and the bib library [here](https://slcladal.github.io/bibliography.bib). You will find very helpful and detailed tutorials on how to perform analyses and visualize eye-tracking data using `eyetrackeR` [here](http://www.eyetracking-r.com/).

## Preparation

```{r install, eval = F, warning=F, message=F}
install.packages(c("tidyverse", "eyetrackingR", "data.table", "itsadug", "sjPlot", "lme4")
```

```{r setup, warning=F, message=F}
# set options
options(stringsAsFactors = F)          # no automatic data transformation
options("scipen" = 100, "digits" = 10) # suppress math annotation
# load packages
library(tidyverse)
library(eyetrackingR)
library(data.table)
library(itsadug)
library(lme4)
library(sjPlot)
```

```{r setup2, echo=F, message=FALSE, warning=FALSE}
# load packages for website
library(knitr) 
library(kableExtra) 
library(DT)
```

# Data processing

## Define paths{-}

```{r}
datapath <- here::here("data/data_exp_50674-v2", "uploads")
masterpath <- here::here("data/data_exp_50674-v2", "data_exp_50674-v2_task-etfm.csv")
```

## Load data{-}

```{r list, warning=F, message=F}
# get list of spreadsheets
fls <- list.files(datapath, full.names = T)
fls <- fls[2:length(fls)]
# create list of tables
datls <- lapply(fls, function(x){
  name <- x
  x <- readxl::read_xlsx(x) %>%
    dplyr::mutate(idname = name) %>%
    dplyr::mutate(trial = stringr::str_remove_all(name, ".*collection-"))  %>%
    dplyr::mutate(trial = stringr::str_remove_all(trial, "-.*"))
  })
# combine spreadsheets into a single data frame
edat <- data.table::rbindlist(datls) %>%
  dplyr::mutate(id = 1:nrow(.)) %>%
  dplyr::mutate(participant_id = factor(participant_id),
                trial = factor(trial))
# extract boundary data
ibs <- edat %>%
  dplyr::select(participant_id, trial, zone_name, zone_x, zone_y,zone_width, zone_height) %>%
  # get rid of superfluous rows
  dplyr::filter(zone_name == "Right"|zone_name == "Left") %>%
  na.omit() %>%
  # define image boundaries
  dplyr::mutate(top = zone_y,
                bottom = zone_y + zone_width,
                left = zone_x,
                right = zone_x + zone_width) %>%
  # remove superfluous columns
  dplyr::select(-zone_x, -zone_y, -zone_width, -zone_height) %>%
  dplyr::mutate(position = tolower(zone_name)) %>%
  tidyr::gather(edge, coordinate, top:right) %>%
  dplyr::mutate(position_edge = paste0(position, "_", edge, "edge")) %>%
  dplyr::select(-zone_name, -position, -edge) %>%
  tidyr::spread(position_edge, coordinate) %>%
  dplyr::group_by(participant_id, trial) %>%
  # create aoi
  dplyr::summarise(left_bottomedge = left_bottomedge,
                   left_leftedge = left_leftedge,
                   left_rightedge = left_rightedge,
                   left_topedge = left_topedge,
                   right_bottomedge = right_bottomedge,
                   right_leftedge = right_leftedge,
                   right_rightedge = right_rightedge,
                   right_topedge = right_topedge)
```

## Process master file{-}

```{r mstr1, warning=F, message=F}
mstr <- read_csv(masterpath) %>%
  # create participant column that matches the participant column in the data 
  dplyr::mutate(participant_id = `Participant Private ID`)
# create reduced master file (no superfluous columns)
mstr_redux <- mstr %>%
  # select columns you need
  dplyr::select(participant_id, `trial number`, condition, target_gender, 
                target_position, `Zone Type`, Response, Correct, target_item) %>%
  # filter unique 
  unique() %>%
  # remove rows containing NA
  na.omit() %>%
  # filter out superfluous rows
  dplyr::filter(`Zone Type` == "response_button_image") %>%
  dplyr::rename(trial = `trial number`) %>%
  dplyr::mutate(trial = as.character(trial)) %>%
  # convert participant_id and trial into factors
  dplyr::mutate(participant_id = factor(participant_id),
                trial = factor(trial))
```

## Join data and masterfile{-}

```{r join1, warning=F, message=F}
edat <- edat %>%
  # remove superfluous columns
  dplyr::select(-`0`, -filename, -spreadsheet_row, -time_stamp, -screen_index, 
                -convergence, -zone_x, -zone_y, -zone_width,
                -zone_height, -zone_x_normalised, -zone_y_normalised, -zone_width_normalised,
                -zone_height_normalised, -idname, 
                
                # WARNING: If you work with normalized values, REPLACE the following 
                # with their non-normalized counterparts!
                
                -x_pred_normalised, -y_pred_normalised) %>%
  # convert participant_id and trial into factors
  dplyr::mutate(participant_id = factor(participant_id),
                trial = factor(trial)) %>%
  # remove rows without gaze information
  dplyr::filter(x_pred != 0,
                y_pred != 0)
edatibs <- left_join(edat, ibs, by = c("participant_id", "trial"))
edatibs <- edatibs %>%
  # filter imprecise data points
  dplyr::filter(face_conf >= .5)
```

## Combining the data and metadata{-}

Now we combine the collected data (`edatibs` = `edat` plus image boundaries `ibs`) with the metadata (the information from the reduced master file `mstr_redux`)

```{r join3, warning=F, message=F}
dat <- dplyr::left_join(edatibs, mstr_redux, by = c("participant_id", "trial"))  %>%
  # determine if participant's gaze was in AOI
  dplyr::mutate(AOI = ifelse(
    # if target is left image
    target_position == "left" &
      y_pred > left_topedge & 
      y_pred < left_bottomedge & 
      x_pred > left_leftedge & 
      x_pred < left_rightedge, 1,
    # if target is right image
    ifelse(target_position == "right" &
             y_pred > right_topedge   & 
             y_pred < right_bottomedge &
             x_pred  > right_leftedge & 
             x_pred <  right_rightedge, 1,
           0)))
```

## Binning{-}

Define time bins (here 100 ms)

```{r bin1, warning=F, message=F}
dat <- dat %>%
  # arrange by participant, trial, and time
  dplyr::arrange(participant_id, trial, time_elapsed) %>%
  # bin times into .2 time bins
  dplyr::mutate(TimeBin = itsadug::timeBins(time_elapsed, 100, pos=0))
```

## Cleaning{-}

```{r clean1, warning=F, message=F}
dat <- dat %>%
  # clean condition (spelling error!)
  dplyr::mutate(condition = dplyr::case_when(condition == "color" ~ "color",
                                             condition == "same" ~ "same",
                                             condition == "different" ~ "different",
                                             condition == "differernt" ~ "different",
                                             TRUE ~ condition)) %>%
  # change correct from 0 vs 1 into correct vs incorrect
  dplyr::mutate(Correct = ifelse(Correct == 1, "Correct",
                                 ifelse(Correct == 0, "Incorrect", Correct)),
                Correct = factor(Correct))
```

## Getting rid of incorrect observations

```{r clean2, warning=F, message=F}
dat <- dat %>%
  dplyr::filter(Correct != "Incorrect")
```

# Inspect processed data


```{r bound1b, echo = F, warning=F, message=F}
dat %>%
  head(10) %>%
  kable(caption = "First 10 rows of dat.") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                            full_width = F)
```


# Saving the data

You can now save the data in your `data` folder, if you like.

```{r save, eval = F, message=F, warning=F}
write.table(dat, here::here("data", "dat.txt"), sep = "\t", row.names = F)
```

To re-load this data, you would have use the following command:

```{r save2, eval = F, message=F, warning=F}
dat <- read.delim(here::here("data", "dat.txt"), sep = "\t")
```

# Citation & Session Info {-}

Schweinberger, Martin. `r format(Sys.time(), '%Y')`. *AcqVA Aurora Guide to Processing Eye-Tracking Data in R*. Tromsø: The Artic University of Norway. url: https://slcladal.github.io/etpro.html (Version `r format(Sys.time(), '%Y.%m.%d')`).

```
@manual{schweinberger`r format(Sys.time(), '%Y')`etpro,
  author = {Schweinberger, Martin},
  title = {AcqVA Aurora Guide to Processing Eye-Tracking Data in R},
  note = {https://slcladal.github.io/eyews.html},
  year = {2021},
  organization = "Arctic University of Norway, AcqVA Aurora Center},
  address = {Tromsø},
  edition = {`r format(Sys.time(), '%Y.%m.%d')`}
}
```


***

[Back to top](#introduction)

***


