---
title: "How to test if including a random effect is warranted?"
author: "Martin Schweinberger"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  bookdown::html_document2: 
    includes:
      in_header: GoogleAnalytics.html
bibliography: bibliography.bib
link-citations: yes
---

# Introduction{-}

This approach is proposed by  @galwey2007introduction[213-214] but it has been contested and may not be conceptually sound (see this discussion on [StackOverflow](https://stats.stackexchange.com/questions/56150/how-can-i-test-whether-a-random-effect-is-significant)) and it is not included in the second edition of the book [see @galwey2014introduction].

# Generate data{-}

Create a constant and use it as a random effect, then create two models: one model with the constant as a random effect and another model with variable as random effect. then use `anova` to compare the models (see below).

```{r}
set.seed(101)
dataset <- expand.grid(id = factor(seq_len(10)), 
                       fac1 = factor(c("A", "B"), levels = c("A", "B")), 
                       trial = seq_len(10))
dataset$value <- rnorm(nrow(dataset), sd = 0.5) +
      with(dataset, rnorm(length(levels(id)), sd = 0.5)[id] +
      ifelse(fac1 == "B", 1.0, 0)) + rnorm(1,.5)
dataset$idconst <- factor(rep(1, each = length(dataset$value)))
# inspect data
head(dataset)
```

# Generate and compare models{-}

```{r modeling, warning=F, message=F}
library(lme4)
fm0 <- lmer(value~fac1+(1|id), 
            data = dataset, 
            REML = FALSE)
fm1 <- lmer(value~fac1+(1|idconst), 
            data = dataset, 
            control=lmerControl(check.nlev.gtr.1="ignore"), 
            REML = FALSE)
# compare models
anova(fm1,fm0)
```

According to the model comparison, the random effect should be included as the fm0 model has the lower AIC as well as BIC.

# References{-}

