---
title: "Descriptive Statistics"
author: "UQ SLC Digital Team"
date: "2 Dezember 2018"
output: html_document
---
# Introduction
This chapte focuses on how to describe and summarize data. To show why this is neccessary, think of the following: you are teaching two different classes in the same school, in the same grade, and at the same level. Both classes take the same exam and, after correcting and grading the exams, someone asks you which class performed better. You could of course say something along the lines of "Well, class A had 5 Bs, 10 Cs, 12 Ds, and 2 Fs while class B had 2 As, 8 Bs, 10 Ds, and 4 Fs" but this answer is not really satisfactory. Descriptive statistics enable you to summarize complex data sets in very few words and using only very basic, and easy to understand, concepts. And this is what we will be dealing with in the following. 

Before delving deeper into what decsritive statistics is, is is useful to have a general idea of how it can be contextualized. Thus, on a more general note, we will be dealing only with one particular subbranch of statistics. Statistics in general can be defined as a branch of mathematics that deals with data collection, organization, analysis, interpretation,  and presentation. As such, statistics can be subdivided into two main areas. *Descriptive statistics* deals with the description of data and their visualization, while  *inferential statistics* deals with data analysis and interpretation (typically, this means testing assumptions about correlations between variables (<www.slcladal.slcladal.github.io>). As stated above, here, we will be dealing with the description of data, especially with *measures of central tendency* and *measures of variability*.

# Measures of central tendency

In linguistics three measures of central tendency are of particular relevance: the *mean*, the *median* and the *mode*. Which measure is appropriate depends on the type of variables and the data distribution one is dealing with and what is the intended aim of the data summary.

## Mean
The mean is used when the data is numeric and normally distributed. The mean is calculated by applying the formula shown below.

\begin{equation}
\bar{x}=\frac{1}{n} \sum_{i=1}^n x_i = \frac{x_{1}+x_{2}+ \dots + x_{n}}{n}
\label{eq:mittel}
\end{equation}

To calculate the mean, sum up all values and divide by the number of valuaes. See the example below for clarification.  Consider, we are interested in the mean length of sentences ina short text, then the first thing we could do would be to list the sentences and their length ina table.


```{r echo = F, results = 'asis'}
library(knitr)
Sentences <- c("Call me Ishmael", "Some years ago -- never mind how long precisely -- having little or no money in my purse, and nothing particular to interest me on shore, I thought I would sail about a little and see the watery part of the world.", "It is a way I have of driving off the spleen, and regulating the circulation.", "Whenever I find myself growing grim about the mouth; whenever it is a damp, drizzly November in my soul; whenever I find myself involuntarily pausing before coffin warehouses, and bringing up the rear of every funeral I meet; and especially whenever my hypos get such an upper hand of me, that it requires a strong moral principle to prevent me from deliberately stepping into the street, and methodically knocking people's hats off--then, I account it high time to get to sea as soon as I can.")
Words <- c(3, 40, 15, 87)
df <- data.frame(Sentences, Words)
kable(df, caption = "Sentences of the first paragraph of Herman Melville's *Moby Dick* and the number of words in each sentence.")
```


To calculate the mean, we need to divide the sum of the number of words per sentence  (145) by the number of sentences (7) (cf. equation (\ref{eq:mittel2})).

\begin{equation}
\frac{3+40+15+87}{4} = \frac{145}{4} = 36.25
\label{eq:mittel2}
\end{equation}

The mean sentences length in our example is 36.25 words

In `R`, the mean is calculated as follows.

```{r echo=TRUE, eval=T, warning=F, message=F}
# create numeric vector
frequencies <- c(3, 40, 15, 87)
# calculate mean
mean(frequencies)
```

## Median

The median is typically used hwen dealing with ordinal variables, i.e. variables that are ordered but not truely numeric. The median is the central value in a de- or increasing ordering of values in a vector. In other words, 50% of values are above and 50% of values are below  the median in a given vector.

If the vector contains an even number of elements, then the two central values are summed up and divided by 2. If the vector contains an uneven number of elements, the median represents the central value.

\begin{equation}
median_{x}=
\begin{cases}
x_{\frac{n+1}{2}} & n\text{ uneven} \\
\frac{1}{2}\bigl(x_{\frac{n}{2}}+x_{\frac{n+1}{2}}\bigr) & n\text{ even}
\end{cases}
\label{eq:median}
\end{equation}

Let's have a look at an example. Consider you are interested in the age stratification of speakers in the private dialogie section of the Irish component of the  *International Corpus of English* (ICE). When tabulating and plotting the age variable you get the following table and graph.


```{r echo = F, results = 'asis'}
library(knitr)
Age <- c("0-10", "19-25", "26-33", "34-41", "42-49", "50+")
Counts <- c(9, 160, 70, 15, 9, 57)
df <- data.frame(Age, Counts)
kable(df, caption = "Number of speakers across age groups in the private dialogie section of the Irish component of the  *International Corpus of English* (ICE)")
```


```{r echo=F, eval=T, warning=F, message=F}
Age <- c("0-18", "19-25", "26-33", "34-41", "42-49", "50+")
Counts <- c(9, 160, 70, 15, 9, 57)
df <- data.frame(Age, Counts)
```

```{r echo=F, eval=T, warning=F, message=F}
barplot(df$Counts,
        ylim = c(0,200),
        ylab = "Frequency",
        xlab = "Age")
mtext(df$Age, 1, 1, at = seq(.7, 6.7, 1.2))
text(seq(.7, 6.7, 1.2), df$Counts +10, df$Counts)
box()
```


```{r fig.width=10, fig.height=8, echo=FALSE, warning=FALSE}
library(png)
ima <- readPNG("resources/median1.png")
plot(c(100, 250), c(300, 450), type = "n", xlab = "", ylab = "", xaxt = "n",yaxt = "n")
lim <- par()
rasterImage(ima, lim$usr[1], lim$usr[3], lim$usr[2], lim$usr[4])
```

The age groups represent an order factor which means that there are catergories with a natural order (here from old to young or vice versa). If we order speakers according to their age from young to old, we geta  vector of length 320. If we then take the central value, i.e. the value of the 160$^{th}$ speaker, we get the median age in the private dialogie section of the Irish component of the  *International Corpus of English* (ICE). 

In R, the median is calculated like below.

```{r median}
# create a vector consisting out of ranks
ranks <- c(rep(1, 9), rep(2, 160), rep(3, 70), rep(4, 15), rep(5, 9), rep(6, 57))
# calculate  median
median(ranks)
```
In our case, the median age is *19-25* because the 160$^{th}$ speaker belongs to the 2$^{nd}$ age group, i.e. the age group with speakers between 19 and 25 years old.  

## Mode

The mode is typically used when dealing with categorical variables and it reports which level of a factor or a categorical variable is the most frequent. Here is an example to illustrate the mode. Consider you are interested where most speakers in the private dialogie section of the Irish component of the *International Corpus of English* are currently residing and you get the fowllowing distribution.

```{r echo = F, results = 'asis'}
library(knitr)
CurrentResidence <- c("Belfast", "Down", "Dublin (city)", "Limerick", "Tipperary")
Speakers <- c(98, 20, 110, 13, 19)
df <- data.frame(CurrentResidence, Speakers)
kable(df, caption = "Number of speakers across couties of current residency in the private dialogie section of the Irish component of the  *International Corpus of English* (ICE)")
```


```{r echo=F, eval=T, warning=F, message=F}
barplot(df$Speakers,
        ylim = c(0,200),
        ylab = "Frequency",
        xlab = "Age")
mtext(df$CurrentResidence, 1, 1, at = seq(.7, 5.5, 1.2))
text(seq(.7, 5.5, 1.2), df$Speakers +10, df$Speakers)
box()
```

```{r fig.width=10, fig.height=8, echo=FALSE, warning=FALSE}
library(png)
ima <- readPNG("resources/modalwert1.png")
plot(c(100, 250), c(300, 450), type = "n", xlab = "", ylab = "", xaxt = "n",yaxt = "n")
lim <- par()
rasterImage(ima, lim$usr[1], lim$usr[3], lim$usr[2], lim$usr[4])
```

The tabulated and visualized data show that the mode is *"Dublin (City)"*, because the largest group (110 speakers) of speakers in the corpus are speakers from the city of Dublin. This means that the "average" speaker in in the private dialogue section of the Irish component of the  *International Corpus of English* (ICE) is from Dublin city.

In \verb!R! the mode is calculated as shown below:

```{r mode}
# create a factor with the current residence of speakers
CurrentResidence <- c(rep("Belfast", 98),         # repeat "Belfast" 98 times
                      rep("Down", 20),            # repeat "Down" 20 times
                      rep("Dublin (city)", 110),  # repeat "Dublin (city)" 110 times
                      rep("Limerick", 13),        # repeat "Limerick" 13 times
                      rep("Tipperary", 19))       # repeat "Tipperary" 19 times
# calculate mode
names(which.max(table(CurrentResidence)))         # extract which level occurs most frequently
```

A word of warning is in order here as only the first(!) maximal value is provided by R even if several categories have the same frequency. 

## Notes on measures of central tendency 
You may ask yourself why there are three different types to calculate the central tendency of a why this is neccessary. To answer these questions consider the following example: Imagine you are interested whether the use of discourse particles differs across two corpora that represent the speech of the same five speakers but different registers. In a  first step, you calculate the relative frequency of discourse particle use and both corpora have a mean of 13.4 particles per 1,000 words. Given the mean, the two corpora do not seem to differ. However, when tabulating and plotting the use of particles by speaker across these two corpora you get the distribution shown in Figure XXX, you will immediately recognize that the mean is not the appropriate measure of te central tendency for the two corpora..  

Man kann sich nun fragen, warum es drei unterschiedliche Maße der zentralen Tendenz gibt und warum diese notwendig sind. Anhand eines fiktiven Beispiels soll dies erläutert werden. Nehmen wir an, sie untersuchen die Häufigkeit des Gebrauchs von Diskurspartikeln in zwei kleinen Korpora, die dieselben Sprecher enthalten, aber in unterschiedlichen Gesprächssituationen. 


Sie berechnen für jeden Sprecher die relative Häufigkeit von Diskurspartikeln pro 1,000 Wörtern pro Korpus und berichten anschließend einen Mittelwert von 13.4 für beide Korpora. Es scheint also so zu sein, dass sich die beiden Korpora hinsichtlich des Gebrauchs von Diskurspartikeln nicht unterscheiden. Wenn man allerdings Tabelle (\ref{tab:zent1}) und Grafik (\ref{fig:zent2}) betrachtet, so fällt sofort auf, dass die Angabe des arithmetischen Mittels keine gute Abbildung der Datenstruktur darstellt.


```{r echo=F, eval=T, warning=F, message=F}
library(knitr)
library(dplyr)
Corpus <- c(rep("C1", 5), 
            rep("C2", 5))
Speaker <- rep(c("A", "B", "C", "D", "E"), 2)  
Frequency <- c(5.2, 11.4, 27.1, 13.7, 9.6, 0.2, 0.0, 1.1, 93.7, 0.4)  
particletable <- data.frame(Corpus, Speaker, Frequency)
A <- c(5.2, 0.2)
B <- c(11.4, 0.0)
C <- c(27.1, 0.1)
D <- c(13.7, 93.7)
E <- c(9.6, 0.4)
barplotdata <- data.frame(A, B, C, D, E)
barplotdata <- as.matrix(barplotdata)
rownames(barplotdata) <- c("Corpus1", "Corpus2")
```

```{r echo=F, eval=T, warning=F, message=F}
kable(particletable, 
      caption = "Relative frequencies of discourse partciles per speaker in two corpora")
```

```{r echo = F, results = 'asis'}
barplot(barplotdata, 
        main = "Use of discourse particles across two corpora", # add title
        xlab="Speaker",                    # add x-axis label
        ylab="Relative Frequency",         # add y-axis label
        ylim = c(0,110),                   # define y-axis range
        col=c("darkblue","red"),           # define colors
        legend = rownames(barplotdata),    # add a legend
        beside=T,                          # stack the bars
        las = 1                            # tick label perpensicular
        )                                  # end barplot
box()                                      # create a box around panel
text(c(1.5,2.5,4.5,5.5,7.5,8.5,10.5,11.5, 13.5,14.5), barplotdata+5, barplotdata) 
```

```{r fig.width=10, fig.height=8, echo=FALSE, warning=FALSE}
library(png)
ima <- readPNG("resources/zent2.png")
plot(c(100, 250), c(300, 450), type = "n", xlab = "", ylab = "", xaxt = "n",yaxt = "n")
lim <- par()
rasterImage(ima, lim$usr[1], lim$usr[3], lim$usr[2], lim$usr[4])
```

Figure XXX shows that the use of discourse particles is distributed rather evenly across speakers in Corpus 1 while 4 out of 5 speakers use almost no discourse particles in Corpus 2 - only speaker D makes extensive use of discourse particles in corpus 2. The high usage frequency of discourse particles by speaker D in corpus 2 causes the mean of corpus 2 to be identical to the mean reported for corpus 1 althpugh the distribution of usage rates differs drastically. This means that reporting the median in addition to the mean can be useful even for numeric variables if the distribution of values is very dissimilar.

To exemplify, we will summarize the distribution of iscourse articles in the two corpora: The use of discourse particles in corpus 1 (mean = 13.7, median = 11.4) is substantailly different from the use of discourse particles in corpus 2 (mean = 13.7, median = 0.4).

It is important to keep in mind that - similar to variable scales - measures of central tendency are downward compatible but not upward compatible. This emans that one may report the median and mode for numeric variables but the mean may only be used for numeric variables but not for categorical variables. In the following, we we have a look at another, more common, way to describe the differences in the distribution between the two corpora, namely, measures of variability. 

# Measures of Variability

Measures of variability provide information about the distribution of values such as whether the data are distributed evenly and do not differ substantially or whether the data are rather heterogeneous and are distributed very unevenly. In the following, we will have a look at the *variance* and the *standard deviation*. 

As before, we will use a practical example to see the usefulness of applying measures of variability. Imagine you dealing with two cities that have the same mean tempertaure per year. However, the variability of temperatures varies differs dramatically between the two cities.

```{r echo = F, results = 'asis'}
library(knitr)
Month <- c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December", "Mean")
CityA <- c(-5, -12, 5, 12, 15, 18, 22, 23, 20, 16, 8, 1, 10.25)  
CityB <- c(7, 7, 8, 9, 10, 13, 15, 15, 13, 11, 8, 7, 10.25)  
temprature <- data.frame(Month, CityA, CityB)
kable(temprature, caption = "Average temperature in two cities by month")
```

```{r echo=F, eval=T, warning=F, message=F}
January <- c(-5.00,  7.00)
February <- c(-12.00,  7.00)
March <- c(5.00,  8.00)
April <- c(12.00,  9.00)
May <- c(15.00, 10.00)
June <- c(18.00, 13.00)
July <- c(22.00, 15.00)
August <- c(23.00, 15.00)
September <- c(20.00, 13.00)
October <- c(16.00, 11.00)
November <- c(8.00,  8.00)
December <- c(1.00,  7.00)
lineplotdata <- data.frame(January, February, March, April, May, June, July, August, September, October, November, December)
lineplotdata <- as.matrix(lineplotdata)
rownames(lineplotdata) <- c("CityA", "City")
```

```{r echo = F, results = 'asis'}
plot(lineplotdata[1,], 
     type = "l", 
     col = "red", 
     axes = F, 
     ylab = "Temperture", 
     xlab = "Month",
     ylim = c(-15, 30),
     lty = 1)
lines(lineplotdata[2,], 
      col = "darkblue",
      lty = 2)
axis(1, 1:12, colnames(lineplotdata), cex = .75, las = 2)
axis(2, seq(-15, 30, 5), seq(-15, 30, 5), las = 2)
box()
legend("topright", 
       lty = c(1,2), 
       col = c("red", "darkblue"), 
       legend = c("City A", "City B"), 
       bty = "n")
```


```{r fig.width=10, fig.height=8, echo=FALSE, warning=FALSE}
library(png)
ima <- readPNG("resources/str1.png")
plot(c(100, 250), c(300, 450), type = "n", xlab = "", ylab = "", xaxt = "n",yaxt = "n")
lim <- par()
rasterImage(ima, lim$usr[1], lim$usr[3], lim$usr[2], lim$usr[4])
```


In the follwoing, we will discuss and calculate different measures of variability for the two cities.

## Range
The range is the simples measure of variabilty and reports the lowest and highest value of a distribution. That is, the range provides minimum and maximum of a vector to show the span of values within a distribution.

In \verb!R!, the range is extracted as shown below.

```{r range}
# create a numeric vector
cityA <- c(-5, -12, 5, 12, 15, 18, 22, 23, 20, 16, 8, 1)
min(cityA); max(cityA) # extract range
```

The lowest temperature value for city A is  -12 degrees Celsius and the highest value is 23 degrees Celsius. The range thus spans from -12 to 23. 

## Interquartile range

The interquartile range (IQR) of a numeric vector informs about how values are distributed. The first IQR ranges from the lowest value to the point that encompasses 25 percent of the data. The second IQR ranges between the maximum value of the first IQR and the point that encompasses 50 percent of the data. The upper limit of the second IQR therefore represents the median value. The third IQR ranges between the upper value of the second IQR and the point that encompases 75 percent of the data. The fourth IQR ranges betwen the upper value of the third IQR and the maximum value. The advantage of the IQR is that the ranges do not need to be even and different IQR values thus provide an approximation of the distribution of the data.

The easiest way to extract the IQR in `R` is to apply the `summary` function to a vector as shown below.

```{r iqr}
summary(cityA) # extract IQR
```

The `summary` function reports that 25 percent od the data fall within -12 and 4 degrees Celsius. The second IQR ranges between 4 and 13.5 degrees Celsius (the median), the third IQR range between 13.5 and 18.5 degrees Celsius, while the fourth IQR ranges between 18.5 and the highest value in the data (23 degrees Celsius). The IQR represent identical width only if the data is distributed normally. 

## Variance

The variance is calculated according to formula XXX (\ref{eq:varianz}). To calculate the variance, each value is substratced from the mean and the result is squared. The squared values are then added and the resulting sum is divided by the number of values minus 1.


$s = \sigma^2 = \frac{1}{n-1} \sum_{i=1}^{n} (x_i - \bar{x})^{2}$


For our example, the variance of temparatures for city A is 123.6591 and 9.477273 for city B.

In `R`, the variance is calculated as shown below:

```{r variance}
sd(cityA)^2
```


## Standard deviation 

The standard deviation (abbreviated with capital $sigma$ $\sigma$) is calculated according to equation  (\ref{eq:sd1}) or, alternatively, according to equation (\ref{eq:sd2}) and it is the square root of the squared variance.

$\sigma = \sqrt{s} = \sqrt{\frac{1}{n-1} \sum_{i=1}^{n} (x_i - \bar{x})^2}$

$\sigma = \sqrt{\frac{ \sum_{i=1}^{n} (x_i - \bar{x})^2}{n-1}}$

For our example, the equaltion (\ref{eq:sd1}) provides a standard devaition of 11.12 for city A and a standard devaition of 3.08 for city B.

In \verb!R!, the standard deviation is calculated as shown below.

```{r sd}
sd(CityA) # calculate standard deviation
```

The standard deviation of tempertaure values of city A is 11.12.

## Measures of variability for non-normal data

Bislang haben wir nicht beachtet, wie die Daten, die wir betrachten normalverteilt sind oder nicht. Im Folgenden werden wir auf Möglichkeiten eingehen nicht normalverteilte Daten zu beschreiben. Im ersten Schritt erstellen wir Beispieldaten und bestimmen anschließend, ob die Daten normalverteilt sind oder nicht. Hierzu nutzen wir den Shapiro-Wilk Test, welcher darüber Auskunft gibt, ob die Zahlenreihe signifikant von der Normalverteilung abweicht oder nicht.


```{r shapiro}
# anmerkungen
k1 <- c(9.2, 11.4, 27.1, 13.7, 9.6)
k2 <- c(0.7, 0.0, 1.1, 68.7, 0.5)
# apply shapiro-wilk test
shapiro.test(k1) # p < .05 = non-normal

shapiro.test(k2) # p < .05 = non-normal
```


The Shapiro-Wilk test shows that the distribution in the data differs significantly froma normal distribution. However, the  Shapiro-Wilk test has been found to be too generous with small samples and too harsh when dealing with large samples. For this erason, it is recommended that the test should always be accompanied by additional eye-balling of the data.



## Quantile

When dealing with non-normal data, the best option to provide a summary of the distribution is to report the quantiles. 



# Excercises

1. Calculate the mean, median, and mode as well as the standrad deviation for the follwoing two vactors.

A: 1, 3, 6, 2, 1, 1, 6, 8, 4, 2, 3, 5, 0, 0, 2, 1, 2, 1, 0 

B: 3, 2, 5, 1, 1, 4, 0, 0, 2, 3, 0, 3, 0, 5, 4, 5, 3, 3, 4 


mean:

median:

mode:

standard deviation:

2. Find a partner and discuss which meaure of central tendency is sppropriate when dealing with grades. Then, find another partner and see whether they hve come to teh same conclusion or discuss why if not. Finally, discuss the advantages and disadvantages of calculateing themmean when dealing with grades.

3. Where are mean, median, and mode when dealing with normal data?

4. Sie erhalten bei drei Untersuchungen mit jeweils zwei Gruppen einmal nahezu identische Steuungen, aber unterschiedliche Mittelwerte, einmal nahezu identische Mittelwerte, aber unterschiedliche Steuungen, und in der letzten Untersuchung sowohl unterschiedliche Mittelwerte und unterschiedliche Streuungen. \\
Wie interpretieren Sie diese drei Ergebnisse?

5. Suchen Sie sich nun einen Partner und diskutieren Sie zusammen, was es auf der konzeptionellen und nicht auf der statistischen Ebene bedeuten könnte, dass zwei Gruppen unterschiedliche Streuungen aufweisen. (Vorsicht: Dies ist nicht so trivial, wie es auf den ersten Blick wirken könnte!)
