---
title: "**Language Technology and Data Analysis Laboratory (LADAL)**"
---


This is the website of the **Language Technology and Data Analysis Laboratory (LADAL)** of the [School of Languages and Cultures](https://languages-cultures.uq.edu.au/) at the [University of Queensland](https://www.uq.edu.au/), Australia.


### About LADAL 

LADAL is there to assist SLC staff and students with respect to data analysis, digital research tools, and other forms of technology. The focus of this site is to provide hands-on tutorials, to provide links to further resources, and short descriptions of digital tools relevant for SLC research.

R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

### Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
