---
title: "**Language Technology and Data Analysis Laboratory (LADAL)**"
---


## Introduction

This is the website of the **Language Technology and Data Analysis Laboratory (LADAL)** of the School of Languages and Cultures at the University of Queensland, Australia.


### About LADAL 

LADAL is there for staff and students of the School of Languages and Cultures to assist with matters of data analysis, digital research tools, and other forms of technology. The focus of this site is to provide hands-on tutorials on using R in rsearch at the School of Languages and Cultures and to provide links and short descriptions of digital tools relevant for SLC research.



data support an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

### Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
