---
title: "Test Notebook: How to convert a R Notebook into a Jupyter Notebook"
output: html_notebook
---

This is a test notebook to see how to convert a R markdown document (a R Notebook) into a Jupyter Notebook.

On the conversion from a Rmd to a ipynb, we can find several answers on StackOverflow suggesting, for example, [this procedure](https://stackoverflow.com/questions/57777009/how-can-i-convert-a-rmd-document-to-a-jupyter-notebook):

>
>After checking the silver bullet seems to be jupytext It allows you to convert from/to mardown, rmarkdown, python, ipynb, ... This can actually allow you a pretty neat workflow
>
>    1. write a simple R script.R script that you can spin into a Rmd document
>    2. use knitr::spin('script.R', knit = FALSE) to transform it to Rmd
>    3. use jupytext --to notebook script.Rmd to create script.ipynb
>    4. share or execute the notebook
>

However, this did not work and we us this approach that is described [here](https://medium.datadriveninvestor.com/transforming-your-rmd-to-ipynb-file-r-markdown-to-python-jupyter-b1306646f50b) .

```{r install, eval = F, message=F, warning=F}
install.packages("here")
install.packages("devtools")
```

```{r installconverter, eval = F, message=F, warning=F}
library(devtools)
devtools::install_github("mkearney/rmd2jupyter")
```

Load packages

```{r pack, message=F, warning=F}
# activate packages
library(here)
library(devtools)
library(rmd2jupyter)
```

Generate list of Rmd files

```{r}
ladalrmds <- list.files(here::here("."), pattern = ".Rmd")
ladalrmds
```




```{r}
# convert 1 notebook
rmd2jupyter::rmd2jupyter("cbs/regex_cb.Rmd")
```

This worked and a ipynb-file is now in our directory.

```{r}
# convert many notebooks
for (f in ladalrmds) rmd2jupyter::rmd2jupyter(f)
```



